# Swiss Freelancer vs. Employee Compensation Calculator

This script calculates the hourly rate a freelancer should charge to maintain the same living standard as a comparably compensated employee in Switzerland. The analysis accounts for differences in:

- Social security contributions (AHV/IV/EO, BVG, ALV)
- Insurance costs
- Non-billable time
- Fixed operational expenses
- Risk compensation for inconsistent work
- Tax implications (VAT)

## Key Assumptions:
- Base annual salary: CHF 110,000
- Working days: 227 days per year (after vacation)
- Freelancer occupancy rate: 80% (accounting for gaps between contracts)
- Three productivity scenarios (low/mid/high) with different non-billable time estimates
- Freelancer works primarily from home (low coworking expense)

```{r}
#| code-fold: true
#| code-summary: "Exact Computation"

# Define base scenario
FullTimeSalary <- 110000  # Annual salary
WeeklyHours <- 42

# Define constants
WorkingDaysPerYear <- 252 - 25  # Work days minus vacation (https://www.arbeitstage.ch/EN/arbeitstage_2024.html)
OccupancyRate <- 0.75           # Expected billable rate for freelancer

# Social security rates (https://www.bsv.admin.ch/bsv/en/home/social-insurance/ahv/contributions.html)
AHV_IV_EO_Rate_Employee <- 0.1050  # 10.5% for employees
AHV_IV_EO_Rate_SelfEmployed <- 0.095  # 9.5% for self-employed
ALV_Rate <- 0.0225              # 2.25% for employees
BVG_Rate_Employee <- 0.09       # 9% for employees
BVG_Rate_SelfEmployed <- 0.13   # 13% for self-employed

# Insurance rates
AccidentInsurance_Employee <- 130.80  # Annual cost (10.90 * 12)
AccidentInsurance_SelfEmployed <- 130.80  # Annual cost
SicknessInsurance_Rate <- 0.015    # 1.5% of salary
LiabilityInsurance_SelfEmployed <- 700  # Professional liability

# Fixed costs
CoworkingSpace <- 200 * 12  # Working primarily from home
HardwareAmortization <- 2000  # Computing server, notebook, monitor, printer amortization
SoftwareSubscriptions <- 1200  # 100 CHF per month
InternetPhone <- 1800  # Internet + phone + mobile internet
MiscOfficeExpenses <- 200 * 12  # Office supplies, utilities
AccountingServices <- 1000  # Basic accounting/tax filing assistance

# Total fixed costs
FixedCosts <- CoworkingSpace + HardwareAmortization + SoftwareSubscriptions + 
              InternetPhone + MiscOfficeExpenses + AccountingServices

# Additional freelancer costs - c(low, mid, high)
MarketingCosts <- c(2000, 3000, 4000)  # Website, networking, client acquisition
EducationCosts <- c(1500, 2000, 3000)  # Courses, conferences, books

# VAT rate in Switzerland
VAT_Rate <- 0.081  # 8.1%

# Calculate daily and annual billable hours - c(low, mid, high) productivity scenarios
HoursPerDay <- WeeklyHours/5 - c(1.9, 2, 2.2)  # Non-billable time: breaks, admin, learning
BillableHoursPerYear <- WorkingDaysPerYear * HoursPerDay

# EMPLOYEE CALCULATIONS
# Employee's contribution to social security
EmployeeSocialSecurity <- (AHV_IV_EO_Rate_Employee / 2) * FullTimeSalary + 
                         (ALV_Rate / 2) * FullTimeSalary + 
                         (BVG_Rate_Employee / 2) * FullTimeSalary

# Employer's contribution (social security + insurances)
EmployerContribution <- (AHV_IV_EO_Rate_Employee / 2) * FullTimeSalary + 
                       (ALV_Rate / 2) * FullTimeSalary + 
                       (BVG_Rate_Employee / 2) * FullTimeSalary + 
                       AccidentInsurance_Employee + 
                       FullTimeSalary * SicknessInsurance_Rate

# Total employer cost
TotalEmployerCost <- FullTimeSalary + EmployerContribution

# Real hourly cost to employer
RealHourlyEmployeeCost <- TotalEmployerCost / BillableHoursPerYear

# FREELANCER CALCULATIONS
# Additional social security costs for self-employed compared to employee
FreelancerSocialSecurityDifference <- AHV_IV_EO_Rate_SelfEmployed * FullTimeSalary + 
                           BVG_Rate_SelfEmployed * FullTimeSalary - EmployeeSocialSecurity

# Insurance costs
FreelancerInsurance <- AccidentInsurance_SelfEmployed + 
                      FullTimeSalary * SicknessInsurance_Rate + 
                      LiabilityInsurance_SelfEmployed

# Risk compensation (need higher base income to account for downtime)
RiskCompensation <- (FullTimeSalary / OccupancyRate) - FullTimeSalary

# Total needed income before VAT
TotalNeededIncome <- FullTimeSalary + FixedCosts + FreelancerSocialSecurityDifference + 
                    FreelancerInsurance + RiskCompensation + 
                    MarketingCosts + EducationCosts

# Freelancer billable hours (affected by occupancy rate)
FreelancerBillableHours <- BillableHoursPerYear * OccupancyRate

# Required hourly rate without and with VAT
FreelancerHourlyRate <- TotalNeededIncome / FreelancerBillableHours
FreelancerHourlyRateWithVAT <- FreelancerHourlyRate * (1 + VAT_Rate)
```


```{r}
#| code-fold: true
#| code-summary: "Results presentation"

# Create detailed results data.frame
results_detailed <- data.frame(
  BillableHoursPerDay = round(HoursPerDay, 1),
  BillableHoursPerYear = round(BillableHoursPerYear),
  TotalEmployerCost = round(TotalEmployerCost),
  RealHourlyEmployeeCost = round(RealHourlyEmployeeCost),
  FixedCosts = FixedCosts,
  MarketingCosts = MarketingCosts,
  EducationCosts = EducationCosts,
  FreelancerSocialSecurityDifference = round(FreelancerSocialSecurityDifference),
  FreelancerInsurance = round(FreelancerInsurance),
  RiskCompensation = round(RiskCompensation),
  TotalNeededIncome = round(TotalNeededIncome),
  FreelancerBillableHours = round(FreelancerBillableHours),
  FreelancerHourlyRate = round(FreelancerHourlyRate),
  FreelancerHourlyRateWithVAT = round(FreelancerHourlyRateWithVAT)
  )
rownames(results_detailed) <- c("Low productivity", "Medium productivity", "High productivity")

# Create summary data.frame with key insights
results_summary <- data.frame(
  Scenario = c("Low productivity", "Medium productivity", "High productivity"),
  BillableHours_Employee = round(BillableHoursPerYear),
  BillableHours_Freelancer = round(FreelancerBillableHours),
  EmployerHourlyCost_CHF = round(RealHourlyEmployeeCost),
  FreelancerRate_CHF = round(FreelancerHourlyRate),
  FreelancerRateWithVAT_CHF = round(FreelancerHourlyRateWithVAT)
  )
rownames(results_summary) <- c("Low", "Medium", "High")

cat("\n## RESULTS\n")
print(t(as.matrix(results_detailed)))

cat("\n## Fixed Costs Breakdown\n")
fixed_costs <- data.frame(
  Category = c("Coworking Space", "Hardware Amortization", "Software Subscriptions", 
               "Internet & Phone", "Office Expenses", "Accounting Services", "Total"),
  Annual_Cost_CHF = c(CoworkingSpace, HardwareAmortization, SoftwareSubscriptions,
                      InternetPhone, MiscOfficeExpenses, AccountingServices, FixedCosts)
)
print(fixed_costs)
```

