---
title: "Taping Study 2"
author: "Markus Kalisch"
format: html
editor: visual
---

## Data set

-   Original file (set to "read-only"):\
    C:\\Users\\kalischm\\teaching\\consultingCases\\MicheleSeiler_KISPI_Aug2023\\Markus_DataFingerfx_P82.xlsx

    ```{r}
    library(readxl)
    require(digest)
    require(tools)
    pth <- "C:/Users/kalischm/teaching/consultingCases/MicheleSeiler_KISPI_Aug2023/"
    pth <- ""
    dataFile <- "Markus_DataFingerfx_P82.xlsx"
    dataFileFull <- paste0(pth, dataFile)
    if (md5sum(file= dataFileFull) == "5db3192e301c646b09ea13198708ba00") {
      cat("md5sum OK - loading file\n")
      datOrig <- read_excel(dataFileFull)
    } else {
      stop("md5sum not correct.")
    }
    ```

Simplify variable names and convert to factor:

```{r}
selectedCols <- c("PatID", 
                  "age",
                  "sex (0=male, 1=female)",
                  "hand (1=right, 2=left)",
                  "finger (2= Zeigefinger, 3=Mittelfinger etc)",
                  "phalanx (1=proximale Phalanx, 2=Mittelphalanx)",
                  "fxtype (1=proximale Fx, 2=Schaftfraktur)",
                  "taping (1=Taping, 0=Gips)",
                  "comfort_th_D5: wie angenehm war Taping bzw Gips? Skala 0-10, 0=am angenehmsten, 10=ganz schlimm",
                  "Analgesie_total: in total, wie lange Schmerzmedikamente (=Angaben Tag 5 PLUS Tag 21)",
                  "comfort_th_D21 wie angenehm war Taping bzw Gips? Skala 0-10, 0=am angenehmsten, 10=ganz schlimm",
                  "secFx_displ_D5: sekundäre Frakturdislokation, dh Rx ap 15° oder mehr oder vorhandender Rotationsfehler",
                  "secFx_displ_D21: Sekundäre Frakturdislokation neu am Tag 21, dh Fxdislokation Frontalebene 15° oder mehr oder neu Rotationsfehler",
                  "TRAM= total range of motion dh Bewegungsumfang sämtlich Fingergelenke in Grad",
                  "Xray_ap_ED (Unfallbild, Röntgen, Frontalebene, Dislokation in Grad)")

colNamesNew <- c("PatID", 
                  "age",
                  "sex",
                  "hand",
                  "finger",
                  "phalanx",
                  "fxtype",
                  "taping",
                  "comfort_th_D5",
                  "Analgesie_total",
                  "comfort_th_D21",
                  "secFx_displ_D5",
                  "secFx_displ_D21",
                  "TRAM",
                  "Xray_ap_ED")
stopifnot(length(colNamesNew) == length(selectedCols))

## select relevant columns
d <- datOrig[,selectedCols]

## simplify col names
colnames(d) <- colNamesNew

## convert selected cols to factor
factorVars <- c(  "sex",
                  "hand",
                  "finger",
                  "phalanx",
                  "fxtype",
                  "taping",
                  "secFx_displ_D5",
                  "secFx_displ_D21")
d$secFx_displ_Total <- as.factor(d$secFx_displ_D5 | d$secFx_displ_D21) ## PRIMARY ENDPOINT
d[factorVars] <- lapply(d[factorVars], as.factor)
# d$sex <- as.factor(d$sex)
# d$hand <- as.factor(d$hand)
# d$finger <- as.factor(d$finger)
# d$phalanx <- as.factor(d$phalanx)
# d$fxtype <- as.factor(d$fxtype)
# d$taping <- as.factor(d$taping) ## TREATMENT
# d$secFx_displ_D5 <- as.factor(d$secFx_displ_D5)
# d$secFx_displ_D21 <- as.factor(d$secFx_displ_D21)
```

## Data update

None

## Data characteristics

```{r}
summary(d)
```

## Checking Ranomization

Michele: "Die Variable ist Xray_ap_ED (Unfallbild, Röntgen, Frontalebene, Dislokation in Grad) für die Frontalenbene und Xray_lat_ED. Bei dieser Variablenerklärung hat sich ein Fehler eingeschlichen, das ist das Bild am Unfalltag und nicht bei der Nachkontrolle (ED= emergency department)"

```{r}
t.test(Xray_ap_ED ~ taping, data = d)$p.value
```

## Patient characteristics

### Number of patients (treatment if coded in col "taping")

```{r}
table(d$taping)
```

### Boys/girls (col "sex")

```{r}
tab_sex <- table(taping = d$taping, sex = d$sex)
tab_sex
fisher.test(tab_sex)$p.value
```

### Age (col "age")

```{r}
summary(d$age[d$taping == "0"])
summary(d$age[d$taping == "1"])
t.test(age~taping, data = d)$p.value
```

### Fracture side (col "hand")

```{r}
tab_hand <- table(taping = d$taping, hand = d$hand)
tab_hand
fisher.test(tab_hand)$p.value
```

### Digit (col "finger")

```{r}
tab_finger <- table(taping = d$taping, finger = d$finger)
tab_finger
chisq.test(tab_finger, simulate.p.value = TRUE, 
           B = 100000)$p.value
```

### Phalanx: proximal/middle (col "phalanx")

```{r}
tab_phalanx <- table(taping = d$taping, phalanx = d$phalanx)
tab_phalanx
fisher.test(tab_phalanx)$p.value
```

### Base/Shaft (col "fxtype")

```{r}
tab_fxtype <- table(taping = d$taping, fxtype = d$fxtype)
tab_fxtype
fisher.test(tab_fxtype)$p.value
```

## Primary endpoint

sec Fx_disp

```{r}
table(disl = d$secFx_displ_Total, 
      taping = d$taping)
```

## GLM --- consider all covariates simultaneously
```{r}
table(broken=d$secFx_displ_Total, finger =d$finger)
# --> only pinky-fingers are displaced (but they are also the ones,
#     most often observed)

fit <- glm(secFx_displ_Total ~ taping + sex + hand + phalanx + fxtype + age, data=d,
  family="binomial")
summary(fit)
confint(fit)

table(broken=d$secFx_displ_Total, finger =d$finger, taping=d$taping)
```

```{r}
require(Epi)
## 'success' = secondary dislocation within first 5 days
aa <- sum( (d$secFx_displ_Total == "TRUE") & (d$taping == "1"), na.rm = TRUE) ## 1; disl & tape
bb <- sum( (d$secFx_displ_Total == "TRUE") & (d$taping == "0"), na.rm = TRUE ) ## 3; disl & cast
cc <- sum( (d$secFx_displ_Total == "FALSE") & (d$taping == "1"), na.rm = TRUE ) ## 55; disl & tape
dd <- sum( (d$secFx_displ_Total == "FALSE") & (d$taping == "0"), na.rm = TRUE ) ## 43; disl & cast
resProp <- ci.pd(aa=aa, bb=bb, cc=cc, dd=dd, print = FALSE) ## for p_tape - p_gips
resProp
ci.pd(table(disl = d$secFx_displ_Total, taping = d$taping))
```

## Secondary endpoints

```{r}
require(mice)
names2 <- c("Analgesie_total", "comfort_th_D5", "comfort_th_D21", "TRAM")
distNA <- apply(d[,names2], 2, function(x) sum(is.na(x))/length(x))
distNA
```

```{r}
par(mfrow = c(1,4))
boxplot(comfort_th_D5 ~ taping, data = d)
boxplot(comfort_th_D21 ~ taping, data = d)
boxplot(Analgesie_total ~ taping, data = d)
boxplot(TRAM ~ taping, data = d)
```

### Comfort after 5 days (VAS) (col "comfort_th_D5")

(small is good)

```{r}
library(coin)
summary(d$comfort_th_D5[d$taping == "0"])
summary(d$comfort_th_D5[d$taping == "1"])
wilcox_test(comfort_th_D5 ~ taping, data = d, conf.int = TRUE)
```

### Comfort after 21 days (VAS) (col "comfort_th_D21")  (small is good)

```{r}
summary(d$comfort_th_D21[d$taping == "0"])
summary(d$comfort_th_D21[d$taping == "1"])

wilcox_test(comfort_th_D21 ~ taping, data = d, conf.int = TRUE)
```

### Analgesia intake (days) (col "Analgesie_total")

```{r}
summary(d$Analgesie_total[d$taping == "0"])
summary(d$Analgesie_total[d$taping == "1"])
cat("Outlier: ",d$PatID[which(d$Analgesie_total == 15)])

wilcox_test(Analgesie_total ~ taping, data = d, conf.int = TRUE)
```

### TRAM (degrees) (col "TRAM") Normalverteilung ok

```{r}
summary(d$TRAM[d$taping == "0"])
summary(d$TRAM[d$taping == "1"])
wilcox_test(TRAM ~ taping, data = d, conf.int = TRUE)
```

### Re-Analysis with multiple imputation and t-tests

```{r}
require(mice)
## imp <- mice(d, m = 50, seed = 43, printFlag = FALSE)
## save(file = "imp.rda", imp)
load("imp.rda")
## Comfort D5
resC5 <- summary(pool(with(imp, lm(comfort_th_D5 ~ taping))), conf.int = TRUE) 
(pC5 <- resC5[2,"p.value"])

## Comfort D21
resC21 <- summary(pool(with(imp, lm(comfort_th_D21 ~ taping))), conf.int = TRUE) 
(pC21 <- resC21[2,"p.value"])

## Analgesie
resAn <- summary(pool(with(imp, lm(Analgesie_total ~ taping))), conf.int = TRUE) 
(pAn <- resAn[2,"p.value"])

## TRAM
resTRAM <- summary(pool(with(imp, lm(TRAM ~ taping))), conf.int = TRUE) 
(pTRAM <- resTRAM[2,"p.value"])


```

### Re-Analysis using MI and wilcox.test (package MKinfer)

"Function mi.wilcox.test also works with package mice."

(https://cran.r-project.org/web/packages/MKinfer/vignettes/MKinfer.html#multiple-imputation-wilcoxon-tests)

```{r}
library(exactRankTests)
library(MKinfer)
mi.wilcox.test(imp, x = "comfort_th_D5", y = "taping")
mi.wilcox.test(imp, x = "comfort_th_D21", y = "taping")
mi.wilcox.test(imp, x = "Analgesie_total", y = "taping")
mi.wilcox.test(imp, x = "TRAM", y = "taping")
```
