---
title: |
  Heteroscedastic Linear Mixed Models with 
  `glmmTMB(dispformula=)` and nlme::lme(weights=varIdent(form=))
author: Lukas Graz  
date: 2024-12-04  
format:
  html:
    toc: true
    toc-location: left
    toc-expand: true
    fig-height: 3
    
# tags:
#   - simulation
#   - power
#   - lmm
---

**Summary**:  
- `glmmTMB` can model heteroscedastic data via the `dispformula` argument (c.f. section @sec-sim-dat-vis-w-disp). But Type I error rate is significantly inflated. P-values are not valid
- `nlme::lme` models give valid p-values when accounting for heteroscedasticity with `weights=varIdent(form=)`. They cannot handly corssed random effects afaik.
- `lmerTest::lmer` has only slightly inflated Type I error rate. I suspect it is less powerful on heteroscedastic data though than `nlme::lme`.

**Example**:  
```
glmmTMB(y ~ trt + (1|id), data = D, dispformula = ~trt)
nlme::lme(y ~ trt, random = ~1|id, weights = varIdent(form = ~1|trt))
```

**Note**: Before modelling heteroscedastic LMM try:  
- fixing heteroscedasticity by transforming the response variable (log, sqrt, etc.)
- Simplify mixed model structure by aggregating data like done in [this post](../lmer-slopes-new)


# Setup Data Simulation
```{r}
#| code-fold: true
library(simr)
library(glmmTMB) 
library(nlme)
library(lmerTest)
library(car)
library(ggplot2)


# Design Matrix
X <- data.frame(
  id  = as.factor(rep(1:40, each = 10)),
  trt = as.factor(rep(c("A", "B", "C", "D"), each = 100))
)

# Make it unbalanced
set.seed(123)
X <- X[-sample(101:300, 130, replace=FALSE),]
# xtabs(~id+trt,X)

# create NULL-lmer-model
model <- makeLmer(y ~ trt + (1|id), 
  fixef=  c(0,0,0,0),
  VarCorr = 9,
  data = X, 
  sigma = 1
  )

# heteroscedastic response
get_hetero_data <- function(n) {
  D <- X
  D$y <- simulate(model, 1)[[1]]
  grpCD <- D$trt %in% c("C", "D")
  D$y[grpCD] <- D$y[grpCD] + rnorm(sum(grpCD), 0, 4)
  return(D)
}

# homoscedastic response
get_homo_data <- function(n) {
  D <- X
  D$y <- simulate(model, 1)[[1]]
  return(D)
}
```


# Visualize Data
## Raw heteroscedastic Data
```{r}
set.seed(123)
Dat_hetero <- get_hetero_data()
ggplot(Dat_hetero, aes(x=trt, y=y, group=id)) + geom_boxplot()
```

## Raw homoscedastic Data
```{r}
ggplot(get_homo_data(), aes(x=trt, y=y, group=id)) + geom_boxplot()
```

## Simulated Data from glmmTMB with Dispersion {#sec-sim-dat-vis-w-disp}
```{r}
Dat_hetero$y_disp <- simulate(
  glmmTMB(y ~ 0+trt + (1|id), data = Dat_hetero, dispformula = ~trt), 1)[[1]]
ggplot(Dat_hetero, aes(x=trt, y=y_disp, group=id)) + geom_boxplot()
```

## Simulated Data from glmmTMB without Dispersion
```{r}
Dat_hetero$y_nodisp <- simulate(
  glmmTMB(y ~ 0+trt + (1|id), data = Dat_hetero                    ), 1)[[1]]
ggplot(Dat_hetero, aes(x=trt, y=y_nodisp, group=id)) + geom_boxplot()
```

# Type I error rate (on heteroscedastic data)
```{r}
#| code-fold: true
#| code-summary: analysis functions code
anova.glmmTMB <- glmmTMB:::Anova.glmmTMB

analysis <- function(nsim = 500, 
                    data_fun = get_hetero_data, 
                    model_fun = glmmTMB, 
                    disp = FALSE){
  # seed <- 123
  PVALs <- parallel::mclapply(1:nsim, \(seed){
      set.seed(seed)
      D <- data_fun()
      if(disp){
        if(identical(model_fun,nlme::lme)){
          fit <- lme(y ~ trt, random = ~1|id, data = D, weights = varIdent(form = ~1|trt))
        } else {
          fit <- model_fun(y ~ trt + (1|id), data = D,  dispformula = ~trt)
        }
      } else {
        if(identical(model_fun,nlme::lme)){
          fit <- lme(y ~ trt, random = ~1|id, data = D)
        } else {
          fit <- model_fun(y ~ trt + (1|id), data = D)
        }
      }
      A <- anova(fit)
      C <- if(identical(model_fun, glmmTMB)) coef(summary(fit))$cond else coef(summary(fit))
      c(anova=A["trt", ncol(A)], `trt(C-A)`=C["trtC", ncol(C)])
  }) |> simplify2array()
  print(knitr::kable(as.data.frame(rbind(
    `fraction(PVALs < .05)` = rowMeans(PVALs < 0.05),
    `binom.test(p = .05)` = apply(PVALs < 0.05, 1, function(x) binom.test(sum(x), nsim, 0.05)$p.value) |> signif(2)
  ))))
  
  PVALS_long <- tidyr::pivot_longer(as.data.frame(t(PVALs)), cols = everything(), names_to = "test", values_to = "PVALs")
  ggplot(PVALS_long, aes(PVALs, group=test, col=test)) + 
    stat_ecdf() + geom_abline(slope=1, intercept=0, col="black") +
    geom_vline(xintercept = 0.05, col="black", linetype="dashed") + scale_x_sqrt() +
     scale_y_sqrt() + ggtitle("ECDF of p-values (sqrt-sqrt scale)")
}
```

## `glmmTMB(y ~ trt + (1|id), dispformula = ~trt)` 
```{r, results="asis"}
analysis(data_fun = get_hetero_data, model_fun = glmmTMB, disp = TRUE)
```

## `glmmTMB(y ~ trt + (1|id))`
```{r, results="asis"}
analysis(data_fun = get_hetero_data, model_fun = glmmTMB)
```

## `lmerTest::lmer(y ~ trt + (1|id))`
```{r, results="asis"}
analysis(data_fun = get_hetero_data, model_fun = lmerTest::lmer)
```

## `lme(y~trt,random=~1|id,weights=varIdent(form=~1|trt))`
```{r, results="asis"}
analysis(data_fun = get_hetero_data, model_fun = nlme::lme, disp = TRUE)
```

## `lme(y ~ trt, random = ~1|id)`
```{r, results="asis"}
analysis(data_fun = get_hetero_data, model_fun = nlme::lme)
```


# Type I error rate (on homoskedastic data)
## `glmmTMB(y ~ trt + (1|id), dispformula = ~trt)` 
```{r, results="asis"}
analysis(data_fun = get_homo_data, model_fun = glmmTMB, disp = TRUE)
```

## `glmmTMB(y ~ trt + (1|id))`
```{r, results="asis"}
analysis(data_fun = get_homo_data, model_fun = glmmTMB)
```

## `lmerTest::lmer(y ~ trt + (1|id))`
```{r, results="asis"}
analysis(data_fun = get_homo_data, model_fun = lmerTest::lmer)
```

## `lme(y~trt,random=~1|id,weights=varIdent(form=~1|trt))`
```{r, results="asis"}
analysis(data_fun = get_homo_data, model_fun = nlme::lme, disp = TRUE)
```

## `lme(y ~ trt, random = ~1|id)`
```{r, results="asis"}
analysis(data_fun = get_homo_data, model_fun = nlme::lme)
```
