---
title: Does the variance of the Random-Effect influence the p-values?
author: Lukas Graz
date: 2023-12-18
permalink: /posts/lmm-effect-of-block-variace
tags: 
  - simulation
  - power
---

We study the classical setup of a randomized complete block design with 2 treatments, 2 replications per block and 5 blocks. We vary the variance of the block effect and study how the p-values of the treatment effect change.

$$
% equatiomatic::extract_eq(fit)
\begin{aligned}
  \operatorname{y}_{i}  &\sim N \left(\alpha_{block(i)} + \beta_{trt(i)}, \sigma^2 \right) \\
    \alpha_{j}  &\sim N \left(\mu_{\alpha_{j}}, {\sigma_{\alpha_{j}}}^2 \right)
    \text{, for block j = 1,} \dots \text{,5}\\
    % contr sum for \beta_{trt(i)}:
    \beta_{1}  &= - \beta_{2}  \quad \text{(contr.sum)}
\end{aligned}
$$

### TLDR
The p-values of the treatment effect are not influenced by the variance of the block effect in our example.

## Simulate Data
```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(gridExtra)
library(lmerTest)
library(multcomp)

options(contrasts = c("contr.sum", "contr.poly"))
```

```{r}
# simulate data for a linear mixed model
# a randomized complete block design with 3 treatments, nblock blocks, nrep replicates per block
# an observation error with standard deviation SD_noise and a block effect with standard deviation SD_block
# a treatment effect with standard deviation SD_trt
get_data <- function(nblock=5, nrep=2, SD_noise=3, SD_block=1, SD_trt=0) {
  data <- expand.grid(
    nrep = 1:nrep,
    trt = as.factor(1:2),
    block = as.factor(1:nblock)
  )
  block_effect <- SD_block * rnorm(nblock)[data$block]
  noise <- SD_noise * rnorm(nrow(data))
  trt_effect <- SD_trt * rnorm(nrow(data))[data$trt]
  data$y <- trt_effect + block_effect + noise
  data
}

```

## Plot Function
```{r}
# function that makes an interactions plot
plot_interactions <- function(data, ...) {
  with(data,
  interaction.plot(
      x.factor = trt,
      trace.factor = block,
      response = y,
      type = "b",
      legend = FALSE,
      ...
    )
  )
}
```


## Perform Analysis
1. Get Data that only differs with respect to SD_block (increasing)  
2. Plot the data for each SD_block 
3. Fit a model with lmer and lm for each SD_block  
4. inspect how the estimates and standard errors of the treatmenteffect change with increasing SD_block  

```{r, fig.height=10, fig.width=10}
SD_blocks <- 0:8
fits <- vector("list", length(SD_blocks)*2)
dim(fits) <- c(length(SD_blocks), 2)
dimnames(fits) <- list(paste0("SD_block=",as.character(SD_blocks)), c("lmer", "lm"))

par(mfrow=c(3,3))
for(i in seq_along(SD_blocks)){
  SD_block <- SD_blocks[i]

  set.seed(2)
  data <- get_data(SD_block=SD_block)
  plot_interactions(data, main=paste0("SD_block=", SD_block))
  
  # lmer
  fits[[i,1]] <- lmer(y ~ trt + (1|block), data=data)
  
  # lm
  fits[[i,2]] <- lm(y ~ trt + block, data=data)
}; par(mfrow=c(1,1))


# apply a function to each element of a list-array
# preserving the dimensions and dimnames
elementwise_apply <- function(L, f){
  L_applied <- lapply(L, f)
  dim(L_applied) <- dim(L)
  dimnames(L_applied) <- dimnames(L)
  L_applied
}

summary_coef_trt1 <- elementwise_apply(fits, function(fit) coef(summary(fit))["trt1",])
```

## Results
Illustration: How `summary_coef_trt1` looks like:
```{r}
summary_coef_trt1["SD_block=5",]

# estimate
elementwise_apply(summary_coef_trt1, function(coef) coef[1])
# standard error
elementwise_apply(summary_coef_trt1, function(coef) coef[2])
# p-value
elementwise_apply(summary_coef_trt1, function(coef) coef[length(coef)])

# random effect variance
sapply(fits[,1], function(fit) VarCorr(fit)$block[1]) |> sqrt() |> signif(2) 
```


# Split Plot

from: [pbkrtest-paper](https://doi.org/10.18637%2Fjss.v059.i09) section 3

Harvesting dates:  
- 1: 2/10
- 2: 21/10  
Plot allocation:
|                    | Block 1        | Block 2        | Block 3        | Time        |
|--------------------|--------------- |--------------- |--------------- |-------------|
| Split-plots (1-15) | h1 h1 h1 h1 h1 | h2 h2 h2 h2 h2 | h1 h1 h1 h1 h1 | Harvesting |
| Sowing (1-15)      | s3 s4 s5 s2 s1 | s3 s2 s4 s5 s1 | s5 s2 s3 s4 s1 | Sowing     |
| Split-plots (16-30)| h2 h2 h2 h2 h2 | h1 h1 h1 h1 h1 | h2 h2 h2 h2 h2 | Harvesting |
| Sowing (16-30)     | s2 s1 s5 s4 s3 | s4 s1 s3 s2 s5 | s1 s4 s3 s2 s5 | Sowing     |

```{r}
library(pbkrtest)
data("beets", package = "pbkrtest")

sug4 <- lmer(sugpct ~ block + sow + harvest + (1 | block:harvest), data = beets)
anova(sug4)

beets$bh <- with(beets, interaction(block, harvest))
splitplot_effect <- getME(sug4, "Z") %*% getME(sug4, "b")
# remove splitplot effect from sugpct
beets$sugpct_without_splitplot <- beets$sugpct - splitplot_effect[,1]
sug_rm <- lmer(sugpct_without_splitplot ~ block + sow + harvest + (1 | block:harvest), data = beets)


# SPE = Split Plot Effect
effect_multipliers <- c(0, 2^(-2:6))
names(effect_multipliers) <- paste0("SPE=", effect_multipliers)
sugpct_with_x_SPEs <- lapply(effect_multipliers, function(effect_multiplier) {
  beets$sugpct_without_splitplot + effect_multiplier * splitplot_effect[,1]
})

# plot the data
plot_list <- list() # Create an empty list to store the plots
# Generate the plots and store them in the list
for(i in seq_along(sugpct_with_x_SPEs)){
  beets$y <- sugpct_with_x_SPEs[[i]]
  p <- ggplot(beets, aes(x = bh, y = y)) + 
  geom_boxplot() + 
  geom_jitter() +
  ggtitle(names(effect_multipliers)[i])
  plot_list[[i]] <- p
}
# Arrange the plots in a 4x2 grid
do.call("grid.arrange", c(plot_list, ncol = 2, nrow = 5))


# for every entry in sugpct_with_x_SPE fit a mixed model and extract the p-value for the effect of harvest
sapply(sugpct_with_x_SPEs, function(sugpct_x_SPE) {
  fit <- lmer(sugpct_x_SPE ~ block + sow + harvest + (1 | block:harvest), data = beets)
  c(
    harvest = anova(fit)["harvest", "Pr(>F)"],
    sow = anova(fit)["sow", "Pr(>F)"]
  )
})

# analog but using the lm function
sapply(sugpct_with_x_SPEs, function(sugpct_x_SPE) {
  lm(sugpct_x_SPE ~ 0 + bh + sow, data = beets) |> 
    glht(linfct = matrix(c(1,1,1,-1,-1,-1,0,0,0,0)/6, nrow = 1)) |> 
    summary() |> 
    coef()
}) #==> stays the same

```
