---
title: One-sided tests for Linear Mixed Models
author: Lukas Graz
date: 2024-11-29
---


1. Simulate (and show) some Data
2. fit linear mixed model
3. anova and two-sided t-test
4. one-sided t-test

<!-- 
# simulate data
Simulate data for 2 groups, 3 repetition,  5 cages per repetition, 2 mice of each group per cage
```{r}
#| code-fold: true
group_effect <- 1

set.seed(321)
D <- data.frame(
  repetition = rep(as.factor(1:3), each=20),
  cage = factor(rep(1:15, each=4)),
  group = rep(c("A", "B"), each=2, times=15)
)

# cage effect
cage_effect <- data.frame(
  cage = unique(D$cage),
  cage_effect = rnorm(length(unique(D$cage)), sd=0.4) |> round(2)
)

D <- merge(cage_effect, D, by="cage", sort=FALSE)
D$V13 <- rnorm(nrow(D), mean=10, sd=2) + D$cage_effect + 
  ifelse(D$group == "B", group_effect, 0)
D$cage_effect <- NULL

knitr::kable(D)
``` 
-->

# Load Data
```{r}
# read 20241202_HT_MC38-tumor volume_last day.xlsx file
library(readxl)
try(setwd("./posts/lmm-one-sided-t-tests"), silent=TRUE)

D <- read_excel("20241202_HT_MC38-tumor volume_last day.xlsx")
D$diameter <- D$V13 ^ (1/3) 
knitr::kable(D)
```
Since Volume is heavily right-skewed, we will use the cubic root of the volume as the response variable. The resulting variable approximately resembles the diameter. 

# fit linear mixed model
```{r}
#| message: false
#| warning: false
library(lmerTest)
D$group <- factor(D$group)

# if we only analyze the first repetition:
# fit <- lmer(diameter ~ group +        (1|cage), data=D)

# otherwise:
fit <- lmer(diameter ~ group + repetition + (1|cage), data=D)
```

# anova and two-sided t-test
```{r}
anova(fit)
summary(fit) # look at the p-value of groupB
```

So groupB is 0.93 bigger than groupA. The p-value is 0.058.

# one-sided t-test
see [this stackoverflow post](https://stats.stackexchange.com/questions/325354/if-and-how-to-use-one-tailed-testing-in-multiple-regression)
```{r}
#| result: asis
one_sided_ttest <- function(fit, which_term="group", greater_than_reference=TRUE){
  COEFS <- coef(summary(fit))
  stopifnot(any(grepl(which_term, names(fixef(fit)))))
  which_term <- grep(which_term, rownames(COEFS))
  t_val <- COEFS[which_term, "t value"]
  df <- COEFS[which_term, "df"]
  p_val <- pt(t_val, df, lower.tail=!greater_than_reference)
  cat("p-value for one-sided t-test for the alternative hypothethis:\n", 
    levels(D$group)[1], if(greater_than_reference) "<" else ">", levels(D$group)[2], ":", p_val, "\n")
  invisible(p_val)
}

# PVALUE: one-sided t-test for groupB > groupA
one_sided_ttest(fit, greater_than_reference=TRUE)

# PVALUE: one-sided t-test for groupB < groupA
one_sided_ttest(fit, greater_than_reference=FALSE)
```


The one-sided p-value corresponds to half the two-sided p-value, if and only if the reference group is the one supposed to be smaller. The reference group is a term in R, and is (unless specified) the groupname that comes first in the alphabet. In this case, the reference group is groupA. So the one-sided p-value for groupB > groupA is 0.029 which also equals 0.058/2.

# Group Sequential Design
The consultancy case that has motivated this post, is doing a group sequential design. The idea is to stop the experiment early if the effect is very strong. The interim analysis is done after each repetition. The stopping rule is to stop if the one-sided p-value is smaller than 0,0221 (using the correction method described in Pocock (1977)).

```{r}
# model with only the first repetition
fit1 <- lmer(diameter ~ group + (1|cage), 
             data=D, subset= repetition==1)
one_sided_ttest(fit1, greater_than_reference=TRUE)

# model with the first two repetitions
fit2 <- lmer(diameter ~ group + repetition + (1|cage), 
             data=D, subset= repetition %in% 1:2)
one_sided_ttest(fit2, greater_than_reference=TRUE)

# model with all three repetitions
fit3 <- lmer(diameter ~ group + repetition + (1|cage), 
             data=D, subset= repetition %in% 1:3)
one_sided_ttest(fit3, greater_than_reference=TRUE)
```
<!-- The p-values are always greater than 0.0221, so we cannot reject the null hypothesis even after 3 repetitions. -->
The first p-value is smaller than 0.0221, so we could have stopped the experiment after the first repetition. 
