---
title: "Testing Slopes in a LMM -- Should we Simplify?"
author: Lukas Graz
date: 2023-05-31
output:
  pdf_document:
    df_print: kable
    toc: true
    toc_depth: 2
    number_sections: true
    fig_width: 6.5
    fig_height: 2
    fig_caption: true
    keep_tex: false
    keep_md: true
---
```{r setup, cache=FALSE, echo=FALSE} 
knitr::opts_chunk$set(cache = TRUE, autodep = TRUE)
set.seed(123)
``` 

# TLDR
Given a dataset as shown in the last plot, we consider the following analyses:  
1. For each subject extract the slope and perform a simple t-test on the slopes  
2. Fit `lmer(salery ~ slope + (1|subject))` and test if $\beta_{slope} =0$  
3.  Fit `lmer(salery ~ slope + (slope|subject))` and test if $\beta_{slope} =0$  
**Then:** 1 & 3 are valid (equally powerful) approaches and 2 is only valid if there is no individual slope.

# Simulate Data

Salary of subject $s$ at time $t$:
$$
salary_{t}^{(s)} = t \cdot (slope + d_s) + intercept_s + \varepsilon_{t,s}
$$
- $d_s \sim \mathcal N(0, subjSlopeSD^2)$  
- $intercept_s \sim \mathcal N(0, subjSD^2)$  
- $\varepsilon_{t,s} \sim \mathcal N(0, obsSD^2)$  

```{r}
library(mcreplicate)
library(ggplot2)
library(lmerTest)

#' `nsub`: how many subjects (default: 6)
#' `nyears`: how many years(default: 10)
#' `obsSD`: standard deviation of noise (observation-level) (default: 15)
#' `subjSD`: standard deviation of individual effect (default: 4)
#' `slope`: shared increase of income per year (default: 5)
#' `subjSlopeSD`: subject specific standard deviation from slope (default: 2)
get_data <- function(nsub=6, nyears=10, 
                     obsSD=15, subjSD=4, 
                     slope=5, subjSlopeSD=2){
  subj_intercept <- rep(rnorm(nsub, 0, subjSD), each=nyears)
  subj_slope <- rep(rnorm(nsub, slope, subjSlopeSD), each=nyears)
  data.frame(
    subject = as.factor(rep(1:nsub, each = nyears)),
    year = rep(1:nyears, times = nsub),
    salary =  subj_intercept +  # subject effect
              subj_slope*(1:nyears) + # individual slope
              rnorm(nyears*nsub, 0, obsSD) # obsSD
  )
}
```
 
# Plot Data
```{r}
plot_data <- function(main, ...){
  ggplot(get_data(...), aes(x = year, y = salary, group = subject, color = subject)) +
    geom_line() +
    geom_point() +
    labs(x = "Year", y = "Salary") +
    theme_minimal() +
    ggtitle(main)
}


```
```{r}
plot_data("Noisless, equal subjects", obsSD=0, subjSD=0, subjSlopeSD=0)
plot_data("Noisless, equal slopes", obsSD=0, subjSlopeSD=0)
plot_data("Noisless", obsSD=0)
plot_data("General case")
```


# Analysis Methods
```{r}
slopeTTest <- function(data){
  fits <- lmList(salary ~ year | subject, data)
  slopes <- coef(fits)[,"year"]
  t_test <- t.test(slopes, mu = 0)
  t_test$p.value
}

lmmRandItcpt <- function(data){
  lmm <- lmer(salary ~ year + (1|subject), data = data)
    summary(lmm)$coefficients[2, "Pr(>|t|)"]
}

lmmRandSlope <- function(data){
  lmm <- lmer(salary ~ year + (year|subject), data = data)
    summary(lmm)$coefficients[2, "Pr(>|t|)"]
}

p_values <- function(...){
  data <- get_data(...)
  c(
    slopeTTest = slopeTTest(data),
    lmmRandItcpt = lmmRandItcpt(data),
    lmmRandSlope = lmmRandSlope(data)
  ) # if you cange the amount of arguments, change also the object "Power"
}

get_power <- function(...){
  args <- list(...)
  PVALS <- as.data.frame(t(
  mc_replicate(2000, do.call(p_values, args))
  ))
  # POWER:
  sapply(lapply(PVALS, function(x) x<0.05), mean)
}
```

# Power Calculations
```{r}
set.seed(123)
ARGS <- as.data.frame(rbind(
  # NULL:
  expand.grid( 
    nsub=6, 
    nyears=10, 
    obsSD=5, 
    subjSD=4, 
    slope=0, 
    subjSlopeSD=c(0,4)),
  # Change standard deviations and effects:
  expand.grid(
    nsub=6, 
    nyears=10, 
    obsSD=c(5,15), 
    subjSD=c(4, 12), 
    slope=c(2,4), 
    subjSlopeSD=c(1,4)),
  # Change samplesize (allocation):
  expand.grid(
    nsub=c(6,10,20), 
    nyears=c(10,20), 
    obsSD=5, 
    subjSD=4, 
    slope=2, 
    subjSlopeSD=1)[-1,]
))
rownames(ARGS) <- NULL

Power <- matrix(NA, nrow=nrow(ARGS), ncol=3)
colnames(Power) <- c("slopeTTest", 
  "lmmRandItcpt", "lmmRandSlope")

for (i in 1:nrow(ARGS)){
  args <- as.list(ARGS[i,])
  Power[i,] <- do.call(get_power, args)
}
```
# Show Results:
```{r, size="tiny"}
as.data.frame(cbind(ARGS, Power))
```
```{r, include=FALSE}
# Perform the two-sample proportion test
result <- prop.test(x = c(4000, 4300), n = c(10000,10000))

# Print the test result
result
result$p.value
```
